---
title: "Gaurav CS5200 Summer2023 Build Triggers"
output: html_notebook
---



```{r}
library(RSQLite)

dbfile = "mediaDB.db"

con <- dbConnect(RSQLite::SQLite(), dbfile)
```



#Q1 Adding Column
```{sql connection=con}

-- Add the new column. It will initially contain NULLs.
ALTER TABLE invoices ADD COLUMN Calculatedtotal NUMERIC(10,2);

```


```{sql connection=con}

-- Copy values from 'Total' to 'Calculatedtotal'.
UPDATE invoices SET Calculatedtotal = Total;

```

#Q2 Check Positive Trigger

```{r}
# Drop the trigger if it exists 
query <- "
DROP TRIGGER IF EXISTS check_positive_calculatedtotal_insert;
"
dbExecute(con, query)

# Trigger for INSERT operation
query <- "
CREATE TRIGGER check_positive_calculatedtotal_insert
BEFORE INSERT ON invoices
FOR EACH ROW
WHEN NEW.Calculatedtotal < 0
BEGIN
    SELECT RAISE(FAIL, 'Calculatedtotal must be positive.');
END;
"
dbExecute(con, query)


# Drop the trigger if it exists
query <- "
DROP TRIGGER IF EXISTS check_positive_calculatedtotal_update;
"
dbExecute(con, query)

# Trigger for UPDATE operation
query <- "
CREATE TRIGGER check_positive_calculatedtotal_update
BEFORE UPDATE ON invoices
FOR EACH ROW
WHEN NEW.Calculatedtotal < 0
BEGIN
    SELECT RAISE(FAIL, 'Calculatedtotal must be positive.');
END;
"
dbExecute(con, query)


```




#Q3 TriggerOnInsert



```{r}
# Create the after insert trigger
query <- "
DROP TRIGGER IF EXISTS update_invoice_total_after_insert;
"
dbExecute(con, query)


query <- "
CREATE TRIGGER update_invoice_total_after_insert
AFTER INSERT ON invoice_items
BEGIN
  UPDATE invoices
  SET Total = (
    SELECT SUM(UnitPrice * Quantity)
    FROM invoice_items
    WHERE InvoiceId = NEW.InvoiceId
  )
  WHERE InvoiceId = NEW.InvoiceId;
END;
"
dbExecute(con, query)
```





#Q4 TriggerOnUpdate

```{r}
# Create the after update trigger

query <- "
DROP TRIGGER IF EXISTS update_invoice_total_after_update;
"
dbExecute(con, query)

query <- "
CREATE TRIGGER update_invoice_total_after_update
AFTER UPDATE ON invoice_items
BEGIN
  UPDATE invoices
  SET Total = (
    SELECT SUM(UnitPrice * Quantity)
    FROM invoice_items
    WHERE InvoiceId = NEW.InvoiceId
  )
  WHERE InvoiceId = NEW.InvoiceId;
END;
"
dbExecute(con, query)
```




#Q4 TriggerOnDelete

```{r}
# Create the after delete trigger

query <- "
DROP TRIGGER IF EXISTS update_invoice_total_after_delete;
"
dbExecute(con, query)

query <- "
CREATE TRIGGER update_invoice_total_after_delete
AFTER DELETE ON invoice_items
BEGIN
  UPDATE invoices
  SET Total = (
    SELECT SUM(UnitPrice * Quantity)
    FROM invoice_items
    WHERE InvoiceId = OLD.InvoiceId
  )
  WHERE InvoiceId = OLD.InvoiceId;
END;
"
dbExecute(con, query)

```




#Q5 Testing Triggers


#Inserting new invoice items and checking the Total
```{r}
# Check the total of a certain invoice before any operations
invoice_id <- 1  # Replace with a valid invoice ID in your database
query <- sprintf("SELECT Total FROM invoices WHERE InvoiceId = %d;", invoice_id)
print(dbGetQuery(con, query))
# 
# # Insert a new invoice item and check the total again

query <- sprintf("INSERT INTO invoice_items (InvoiceId, TrackId, UnitPrice, Quantity) VALUES (%d, 1, 1.99, 2);", invoice_id)
dbExecute(con, query)
query <- sprintf("SELECT Total FROM invoices WHERE InvoiceId = %d;", invoice_id)
print(dbGetQuery(con, query))
```




#Updating the invoice_items and checking total

```{r}
# Update the invoice item and check the total again

query <- "UPDATE invoice_items SET Quantity = 3 WHERE InvoiceId = 1 AND TrackId = 1;"
dbExecute(con, query)
query <- sprintf("SELECT Total FROM invoices WHERE InvoiceId = %d;", invoice_id)
print(dbGetQuery(con, query))
```



#Deleting the Invoice item and cheking the total

```{r}
# Delete the invoice item and check the total again

query <- "DELETE FROM invoice_items WHERE InvoiceId = 1 AND TrackId = 1;"
dbExecute(con, query)
query <- sprintf("SELECT Total FROM invoices WHERE InvoiceId = %d;", invoice_id)
print(dbGetQuery(con, query))
```



#Inserting the negative Total value and checking the Trigger
```{r}
query <- "
INSERT INTO invoices (InvoiceId, CustomerId, InvoiceDate, Calculatedtotal) 
VALUES (413, 44, '2023-06-14', -10);
"
dbExecute(con, query)
```


#Updating the invoice with negative Total and checking Trigger

```{r}
# Update the invoice Total and check the Trigger

query <- "UPDATE invoices SET Calculatedtotal = -10 WHERE InvoiceId = 1;"
dbExecute(con, query)
```



```{r}
dbDisconnect(con)
```


