---
title: "Practicum I CS5200"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
Author: Gaurav
Date: Summer Full 2023
---



## Connect to Database

```{r load_packages}
# Install and load the 'RMySQL' package
install.packages("RMySQL")
library(RMySQL)

# install.packages('plyr', repos = "http://cran.us.r-project.org")
options(repos = "https://cloud.r-project.org/")
install.packages("plyr")
```


```{r Connect_Database}
# Create a connection object
con <- dbConnect(
  MySQL(),
  user = "sql8628033",
  password = "qSWxNVaMtl",
  dbname = "sql8628033",
  host = "sql8.freemysqlhosting.net"
  # user = "sql12629017",
  # password = "cFgCnCYFcX",
  # dbname = "sql12629017",
  # host = "sql12.freemysqlhosting.net"
)

# Check if the connection was successful
if (dbIsValid(con)) {
  cat("Connection to MySQL database established successfully!")
} else {
  cat("Failed to connect to the MySQL database.")
}
```



## Create Database

```{sql strikes_table_drop, connection=con}
DROP TABLE IF EXISTS strikes;
```


```{sql flights_table_drop, connection=con}
-- Drop the 'flights' table if it exists
DROP TABLE IF EXISTS flights;
```



```{sql conditions_table_drop, connection=con}
-- Drop the 'conditions' table if it exists
DROP TABLE IF EXISTS conditions;
```



```{sql airports_table_drop,connection=con}
-- Drop the 'airports' table if it exists
DROP TABLE IF EXISTS airports;
```




### Create airports table

```{sql airports_table_create,connection=con}
-- Create the 'airports' table

CREATE TABLE airports (
  aid INT NOT NULL AUTO_INCREMENT,
  airportState TEXT,
  airportCode TEXT,
  PRIMARY KEY (aid)
);

```


### Create flights table

```{sql flights_table_create, connection=con}
# Create the 'flights' table

CREATE TABLE flights (
  fid INT NOT NULL AUTO_INCREMENT,
  date DATE,
  origin INT,
  airline TEXT,
  aircraft TEXT,
  altitude INT CHECK (altitude > 0),
  heavy BOOLEAN,
  PRIMARY KEY (fid),
  FOREIGN KEY (origin) REFERENCES airports(aid)
);

```


### Create conditions table

```{sql conditions_table_create, connection=con}
-- Create the 'conditions' lookup table

CREATE TABLE conditions (
  cid INT NOT NULL AUTO_INCREMENT,
  sky_condition TEXT,
  explanation TEXT,
  PRIMARY KEY (cid)
);

```


### Create strikes table

```{sql strikes_table_create, connection=con}
-- Create the 'strikes' table

CREATE TABLE strikes (
  sid INT NOT NULL AUTO_INCREMENT,
  fid INT,
  numbirds INT,
  impact TEXT,
  damage BOOLEAN,
  altitude INT CHECK (altitude > 0),
  conditions INT,
  PRIMARY KEY (sid),
  FOREIGN KEY (fid) REFERENCES flights(fid),
  FOREIGN KEY (conditions) REFERENCES conditions(cid)
);

```
```{r}
tables <- dbGetQuery(con, "SHOW TABLES;")

  # Loop through each table and retrieve its schema
  for (table_name in tables[,1]) {
    query <- paste("SHOW COLUMNS FROM", table_name, ";", sep = " ")
    schema <- dbGetQuery(con, query)
    print(schema)
  }
```





## Loading CSV file
```{r csv_load}
# Load the bird strikes CSV file
bds.raw <- read.csv("BirdStrikesData-V2.csv")
#bds.subset <- bds.raw
```

 
### Create a subset of loaded data

```{r data_subset}
#  # Set the seed for reproducibility
set.seed(123)

# Create a subset of the data
subset_size <- 500  # Adjust the subset size as needed
bds.subset <- bds.raw[sample(nrow(bds.raw), subset_size), ]

# View the subsetted data
head(bds.subset)

```


## Populate Airport Table

```{r populate_airports}
# Determine the unique airport states from the dataframe
unique_states <- unique(bds.subset$origin)

# Create a dataframe to store the airport data
airports_data <- data.frame(airportState = unique_states, airportCode = character(length(unique_states)), stringsAsFactors = FALSE)

head(airports_data)

# Insert the "unknown" airport into the dataframe (for cases where airport information is missing)
airports_data <- rbind(airports_data, data.frame(airportState = "unknown", airportCode = "", stringsAsFactors = FALSE))



# Insert the airport data into the 'airports' table
for (i in 1:nrow(airports_data)) {
  query <- paste0("INSERT INTO airports (airportState, airportCode) VALUES ('", airports_data[i, "airportState"], "', '", airports_data[i, "airportCode"], "')")
  dbExecute(con, query)
}


```

### Print airports table

```{sql print_airports, connection=con}
SELECT * FROM airports LIMIT 100;
```


## Populate the flights table

```{r populate_flights}
# Create a dataframe to store the flight data
flights_data <- data.frame(date = as.Date(bds.subset$flight_date, format = "%m/%d/%Y %H:%M")
, origin = bds.subset$origin, airline = bds.subset$airline, aircraft = bds.subset$model, altitude = bds.subset$altitude_ft, heavy = ifelse(bds.subset$heavy_flag == "Yes", TRUE, FALSE))

print(head(flights_data, 20))

# Replace missing airport values with the "unknown" airport code
flights_data$origin[is.na(flights_data$origin)] <- "unknown"


#Insert the flight data into the 'flights' table
for (i in 1:nrow(flights_data)) {
  date <- ifelse(is.na(flights_data[i, "date"]), "NULL", paste0("'", flights_data[i, "date"], "'"))
  
  
  origin <- paste0("(SELECT aid FROM airports WHERE airportState = '", flights_data[i, "origin"], "' LIMIT 1)")
  

  # Start by copying the airline column to a new vector
airline <- flights_data[i, "airline"]

# Where the value is NA or an empty string, replace with "unknown"
airline[is.na(airline) | airline == ""] <- "unknown"

# Where the value is not "NULL", wrap in single quotes
airline[airline != "NULL"] <- paste0("'", airline[airline != "NULL"], "'")



# For aircraft
aircraft <- flights_data[i, "aircraft"]
aircraft[is.na(aircraft) | aircraft == ""] <- "unknown"
aircraft[aircraft != "NULL"] <- paste0("'", aircraft[aircraft != "NULL"], "'")

# For altitude
altitude <- flights_data[i, "altitude"]
altitude[is.na(altitude) | altitude == ""] <- 0
altitude[altitude != "NULL"] <- gsub(",", "", altitude[altitude != "NULL"])


  heavy <- as.integer(flights_data[i, "heavy"])

  
  query <- paste0("INSERT INTO flights (date, origin, airline, aircraft, altitude, heavy) SELECT ", date, ", (", origin, "), ", airline, ", ", aircraft, ", ", altitude, ", ", heavy)

  # dbExecute(con, query)

  tryCatch({
    dbExecute(con, query)
  }, error = function(e) {
    print(paste("Error on row", i, ":"))
    print(flights_data[i,])
  })

}
 


```

### Print flights table

```{sql print_flights, connection=con}
SELECT * FROM flights LIMIT 100;
```


## Populate the conditions table

```{r populate_conditions}
# Determine the unique sky conditions from the dataframe
unique_conditions <- unique(bds.subset$sky_conditions)

# Create a dataframe to store the conditions data
conditions_data <- data.frame(sky_condition = unique_conditions, explanation = "")

# Insert the conditions data into the 'conditions' table
for (i in 1:nrow(conditions_data)) {
  query <- paste0("INSERT INTO conditions (sky_condition, explanation) VALUES ('", conditions_data[i, "sky_condition"], "', '')")
  dbExecute(con, query)
}

```


### Print conditions table

```{sql print_conditions, connection=con}
SELECT * FROM conditions;
```


## Populate the strikes table

```{r populate_strikes}

# Create a dataframe to store the strike data

strikes_data <- data.frame(date = as.Date(bds.subset$flight_date, format = "%m/%d/%Y %H:%M"), origin = bds.subset$origin, airline = bds.subset$airline, aircraft = bds.subset$model, numbirds = bds.subset$wildlife_struck, impact = bds.subset$impact, damage = bds.subset$damage, altitude = as.integer(gsub(",", "", bds.subset$altitude)), conditions = bds.subset$sky_conditions)




strikes_data$damage <- as.character(strikes_data$damage)
strikes_data$damage <- ifelse(strikes_data$damage == "No damage", "FALSE", "TRUE")
strikes_data$damage <- as.logical(strikes_data$damage)

# Replace missing or blank values with appropriate defaults
strikes_data$date[is.na(strikes_data$date) | strikes_data$date == ""] <- NULL
strikes_data$airline[is.na(strikes_data$airline) | strikes_data$airline == ""] <- "unknown"
strikes_data$aircraft[is.na(strikes_data$aircraft) | strikes_data$aircraft == ""] <- "unknown"
strikes_data$numbirds[is.na(strikes_data$numbirds)] <- 0
strikes_data$impact[is.na(strikes_data$impact) | strikes_data$impact == ""] <- "unknown"
strikes_data$altitude[is.na(strikes_data$altitude)] <- 0
strikes_data$conditions[is.na(strikes_data$conditions) | strikes_data$conditions == ""] <- "unknown"

# Remove duplicate strikes for the same flight (keeping only the first strike)

strikes_data <- strikes_data[!duplicated(strikes_data[c("date", "airline")]), ]


for (i in 1:nrow(strikes_data)) {
  origin <- paste0("(SELECT aid FROM airports WHERE airportState = '", strikes_data[i, "origin"], "' LIMIT 1)")
  query <- paste0("INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions) SELECT fid, ", strikes_data$numbirds[i], ", '", strikes_data$impact[i], "', ", strikes_data$damage[i], ", ", strikes_data$altitude[i], ", (SELECT cid FROM conditions WHERE sky_condition = '", strikes_data$conditions[i], "') FROM flights WHERE airline = '", strikes_data$airline[i], "' AND aircraft = '", strikes_data$aircraft[i], "' AND origin = ", origin, " AND altitude = ", strikes_data$altitude[i], " LIMIT 1")
  dbExecute(con, query)
}


```

### Print strikes table

```{sql print_strikes, connection=con}
SELECT * FROM strikes LIMIT 200;
```


## Q8  Top 10 states with the greatest number of bird strike incidents

```{sql Q8_query, connection=con}
SELECT a.airportState AS state, COUNT(*) AS num_incidents
FROM strikes s
JOIN flights f ON s.fid = f.fid
JOIN airports a ON f.origin = a.aid
GROUP BY a.airportState
ORDER BY num_incidents DESC
LIMIT 10;

```



### List of airlines by incidents

```{sql airline_vs_incidents, connection=con}
SELECT f.airline AS airline, COUNT(DISTINCT s.sid) AS num_incidents
FROM strikes s
JOIN flights f ON s.fid = f.fid
GROUP BY f.airline
ORDER BY num_incidents DESC;

```

### calculating the average of incidents

```{sql avg_incidents, connection=con}
SELECT AVG(num_incidents) AS average_incidents
FROM (
  SELECT COUNT(DISTINCT s.sid) AS num_incidents
  FROM strikes s
  JOIN flights f ON s.fid = f.fid
  GROUP BY f.airline
) AS subquery;

```


## Q9 Find the airlines that had an above average number bird strike incidents.

```{sql Q9_above_avg_incidents, connection=con}
SELECT f.airline AS airline, COUNT(DISTINCT s.sid) AS num_incidents
FROM strikes s
JOIN flights f ON s.fid = f.fid
GROUP BY f.airline
HAVING COUNT(DISTINCT s.sid) > (SELECT AVG(num_incidents) FROM (
  SELECT COUNT(DISTINCT s.sid) AS num_incidents
  FROM strikes s
  JOIN flights f ON s.fid = f.fid
  GROUP BY f.airline
) AS subquery)
ORDER BY num_incidents DESC;

```

### Months vs number of birds

```{sql month_vs_noOfBirds, connection=con}
SELECT MONTH(f.date) AS month, SUM(s.numbirds) AS total_birds
FROM strikes s
JOIN flights f ON s.fid = f.fid
GROUP BY MONTH(f.date)
ORDER BY MONTH(f.date)

```


## Q10 The (total) number of birds that struck aircraft by month

```{r Q10_month_vs_noOfBirds}
# Execute the SQL query and save the result in a dataframe
result <- dbGetQuery(con, "SELECT MONTH(f.date) AS month, SUM(s.numbirds) AS total_birds
FROM strikes s
JOIN flights f ON s.fid = f.fid
GROUP BY MONTH(f.date)
ORDER BY MONTH(f.date)")

# Display the first six rows of the dataframe
head(result, 6)


```
## Q11 Build a column chart that plots month along the x-axis versus number of birds on the y-axis.

```{r chart_for_Q10}
# Create a data frame with all months
all_months <- data.frame(month = 1:12)

# Merge it with your result data frame
result <- merge(all_months, result, by = "month", all.x = TRUE)


# Replace NA values with 0
result$total_birds[is.na(result$total_birds)] <- 0

# Calculate the maximum value of the total_birds column
max_value <- max(result$total_birds)

# Set a buffer space as a percentage of the maximum value
buffer <- 0.1  # Adjust this value as desired (e.g., 0.1 represents 10%)

# Calculate the adjusted y-axis limits
y_limit <- max_value + (max_value * buffer)


# Convert month to factor to treat it as categorical variable
result$month <- factor(result$month, levels = 1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

# Create a bar plot using the barplot function
barplot(height = result$total_birds, names.arg = result$month,
        xlab = "Month", ylab = "Total Number of Bird Strikes", 
        main = "Total Number of Bird Strikes by Month",
        col = "steelblue", ylim = c(0, y_limit))

# Add a grid
grid()
```

## Q12 Create a stored procedure in MySQL.

```{sql Q12_drop_procedure, connection=con}
DROP PROCEDURE IF EXISTS AddStrike;
```


```{sql Q12_create_procedure, connection=con}

CREATE PROCEDURE AddStrike(
     IN p_flight_date DATE,
    IN p_airline VARCHAR(255),
    IN p_origin_state VARCHAR(255),
    IN p_aircraft VARCHAR(255),
    IN p_altitude INT,
    IN p_numbirds INT,
    IN p_impact VARCHAR(255),
    IN p_damage BOOLEAN,
    IN p_conditions VARCHAR(255),
    IN p_heavy BOOLEAN
)
BEGIN
    DECLARE v_airport_id INT;
    DECLARE v_flight_id INT;
    DECLARE v_conditions_id INT;
    
    -- Check if the origin airport already exists
    SELECT aid INTO v_airport_id FROM airports WHERE airportState = p_origin_state;
    
    -- If the origin airport doesn't exist, insert a new record
    IF v_airport_id IS NULL THEN
        INSERT INTO airports (airportState) VALUES (p_origin_state);
        SET v_airport_id = LAST_INSERT_ID();
    END IF;
  
   
  
    
    -- Check if the flight already exists
    SELECT fid INTO v_flight_id FROM flights WHERE date = p_flight_date AND airline = p_airline AND origin = v_airport_id;
    
    -- If the flight doesn't exist, insert a new record
    IF v_flight_id IS NULL THEN
        INSERT INTO flights (date, origin, airline, aircraft, altitude, heavy) VALUES (p_flight_date, v_airport_id, p_airline, p_aircraft, p_altitude, p_heavy);
        SET v_flight_id = LAST_INSERT_ID();
    END IF;
    
    
    
    
    
    -- Check if the conditions already exist
    SELECT cid INTO v_conditions_id FROM conditions WHERE sky_condition = p_conditions;
    
    -- If the conditions don't exist, insert a new record
    IF v_conditions_id IS NULL THEN
        INSERT INTO conditions (sky_condition) VALUES (p_conditions);
        SET v_conditions_id = LAST_INSERT_ID();
    END IF;
    
    
    
    -- Insert the strike record
    INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions)
    VALUES (v_flight_id, p_numbirds, p_impact, p_damage, p_altitude, v_conditions_id);
    
   -- SELECT * FROM strikes WHERE sid = LAST_INSERT_ID();
    
END;


```



```{r Q12_query}


query <- "CALL AddStrike('2023-06-30', 'Business Airlines', 'New York', 'Boeing 737', 20000, 3, 'Engine damage', TRUE, 'No Cloud', TRUE)"
dbExecute(con, query)

# Get the ID of the last inserted row
last_id <- dbGetQuery(con, "SELECT LAST_INSERT_ID()")[[1]]

# Fetch and print the new row from the strikes table
new_strike <- dbGetQuery(con, paste0("SELECT * FROM strikes WHERE sid = ", last_id))
print(new_strike)


```



```{sql last_10_strikes, connection=con}
SELECT *
FROM strikes
ORDER BY sid DESC
LIMIT 10;

```

```{sql last_10_flights, connection=con}
SELECT *
FROM flights
ORDER BY fid DESC
LIMIT 10;

```


```{sql last_5_conditions, connection=con}
SELECT *
FROM conditions
ORDER BY cid DESC
LIMIT 5;

```



```{r last_10_strikes_r}
strike_results <- dbGetQuery(con, "SELECT *
FROM strikes
ORDER BY sid DESC
LIMIT 10")
print(strike_results)
```



```{r disconnect_database}
dbDisconnect(con)
```


